---
title: "Class9"
author: "Laura Biggs"
format: pdf
---

## Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy?

86.2% of the structures in the PDB are solved by X-ray crystallography while only 6.5% of the structures are visualized by EM.

```{r}
#Read in PDB data
PDB <- read.csv("Data Export Summary.csv", row.names = 1)
print(sapply(PDB, class))

#Must convert character columns into numeric
PDB$X.ray <- as.numeric(gsub(",","",PDB$X.ray))
PDB$X.ray
Xray_sum <- sum(PDB$X.ray)

PDB$EM <- as.numeric(gsub(",","",PDB$EM))
PDB$EM
EM_sum <- sum(PDB$EM)

PDB$Total <- as.numeric(gsub(",","",PDB$Total))
PDB$Total
Total_sum <- sum(PDB$Total)

print(sapply(PDB, class))

#% Xray
Total_Xray <- (Xray_sum/Total_sum) * 100
print(Total_Xray)

#% EM
Total_EM <- (EM_sum/Total_sum) * 100
print(Total_EM)
```

## Q2: What proportion of structures in the PDB are protein?
Most of the structures in the PDB are protein at a proportion of .87, or 87%.

```{r}
#Convert character NMR to numeric
PDB$NMR <- as.numeric(gsub(",","",PDB$NMR))
PDB$NMR
print(sapply(PDB, class))


NMR_protein <- PDB["Protein (only)", "Total"]
NMR_protein

#Proportion
NMR_protein_proportion <- NMR_protein/Total_sum
NMR_protein_proportion
```

## Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?
There are 43,831 HIV-1 protease structures in the PDB.

## HIV-Pr

## Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure? 
Hydrogen is too small (1.9 angstroms) to be captured by the resolution of this X ray crystallography (2 angstroms). Only the oxygens can be visualized as they are large enough.

## Q5: There is a critical "conserved" water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have? 
Water molecule 308 is conserved at the binding site.

## Q6: Generate and save a figure clearly showing the two distinct chains of HIV-protease along with the ligand. You might also consider showing the catalytic residues ASP 25 in each chain (we recommend "Ball & Stick" for these side-chains). Add this figure to your Quarto document. 
![](images/1HSG%20(1).png)


## Bio3D

```{r}
# Load in Bio3D
library(bio3d)

# Read in PDB file
pdb <- read.pdb("1hsg")
pdb
```

## Q7: How many amino acid residues are there in this pdb object? 
There are 198 amino acid residues.

## Q8: Name one of the two non-protein residues? 
The non-protein residues are the water molecules and MK1, the protease inhibitor drug.

## Q9: How many protein chains are in this structure?
There are 2 protein chains, chain A and B of the HIV protease.

Inspect pdb further
```{r}
attributes(pdb)
head(pdb$atom)
```

Package Setup
```{r}
#install.packages("ggrepel")
#install.packages("devtools")
#install.packages("BiocManager)

#BiocManager::install("msa")
#devtools::install_bitbucket("Grantlab/bio3d-view")
```

## Q10. Which of the packages above is found only on BioConductor and not CRAN? 
MSA is a package only available on BioConductor.

## Q11. Which of the above packages is not found on BioConductor or CRAN?: 
bio3d-view is not found on BioConductor or CRAN.

## Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?
True

Search & retrieve ADK structures
```{r}
library(bio3d)
aa <- get.seq("1ake_A")
aa
```

## Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 
There are 214 amino acids in the sequence

Blast query
```{r}
b <- blast.pdb(aa)

#hits <- NULL
#hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')

#Summary of search results
hits <- plot(b)

#'Top hits'
head(hits$pdb.id)
```

Download hits as pdb files
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Align and superimpose structure data
```{r}
# Align related PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

#Vectorize PDB codes for figure axis
ids <- basename.pdb(pdbs$id)

#Draw schematic alignment
plot(pdbs, labels=ids)
```

Annotate PDB structures
```{r}
anno <- pdb.annotate(ids)
unique(anno$source)
```

## PCA
```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

Pairwise RMSD
```{r}
rd <- rmsd(pdbs)

#Hierarchical clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

Alternative ggplot
```{r}
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```
Normal mode analysis
```{r}
modes <- nma(pdbs)
plot(modes, pdbs, col=grps.rd)
```

## Q14: What do you note about this plot? Are the black and colored lines similar or different? Where do you think they differ most and why?
The black and colored lines differ in the degree of fluctuation, where colored lines are generally more dynamic. These lines differ most at the functional regions of the protein, ie: where the conformations change. 