---
title: "HalloweenMiniProject"
author: "Laura Biggs"
format: pdf
---

## Load in candy data
```{r}
candy_file <- 'candy-data.csv'

candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

Q1. How many different candy types are in this dataset?
There are 12 different candy types in this dataset, signified by different columns.
```{r}
dim(candy)
```

Q2. How many fruity candy types are in the dataset?
There are 38 candies that fall into the fruity category.
```{r}
sum(candy$fruity)
```

## Using winpercent

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
I like sour gummy worms, particularly the Trolli Sour Bites. The Trolli candy has a winpercent value of 47.17, meaning that people choose this candy over others less than 50% of the time. 
```{r}
candy["Trolli Sour Bites", ]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
Kit Kat's are popular with a winpercent of 76.7.
```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
Tootsie roll snack bars are less popular with a winpercent of 49.6
```{r}
candy['Tootsie Roll Snack Bars', ]$winpercent
```


## Using the skimr package
```{r}
#install.packages("skimr")
library(skimr)
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
The winpercent variable is scaled differently relative to the other variables, and is scaled by 100 rather than 1.

Q7. What do you think a zero and one represent for the candy$chocolate column?
A zero likely represents a candy that does not have chocolate, or otherwise answers false (=0) to the logical, while a one represents a candy that contains chocolate and answers true (=1) to the logical. 


## Plotting the data
Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?
The distribution isn't quite symmetrical or bell shaped, and is somewhat skewed as the frequency of observations cluster from 30%-50% rather than being evenly distributed around 50%. 

Q10. Is the center of the distribution above or below 50%?
The center of the distribution is below 50%, at 47.8%.
```{r}
median(candy$winpercent)
```

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
Chocolate candy is ranked higher than fruity candy on average.
```{r}
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
mean(chocolate)

fruity <- candy$winpercent[as.logical(candy$fruity)]
mean(fruity)
```

Q12. Is this difference statistically significant?
This difference is statistically significant with a p value of 2.871e-08.
```{r}
t.test(chocolate, fruity)
```

## Overall candy rankings
Q13. What are the five least liked candy types in this set?
```{r}
library(dplyr)
candy %>%
  arrange(winpercent) %>%
  head(5)
```

Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```

Barplots
Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
# with ggplot2
library(ggplot2)


ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

Add color to barplot
```{r}
# Make color vectors
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}
#Add to barplot
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(col=my_cols)
```

Q17. What is the worst ranked chocolate candy?
Sixlets are the worst ranked chocolate candy.

Q18. What is the best ranked fruity candy?
Starburst are the best ranked fruity candy.

## Pricepercent
Plot pricepercent vs winpercent
```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Tootsie roll midgies are the highest ranked, 45.7%, for the lowest cost relative to other candies.
```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```


Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
The 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hersheys Krackel, Hersheys Milk Chocolate. Nik L Nip is the least popular.
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Corelation Structure
Using the corrplot package
```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolate and fruity are anti-correlated.

Q23. Similarly, what two variables are most positively correlated?
Chocolate and bar are most positively correlated.

## PCA
```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

PCA plot
```{r}
# x plots scores, not rotation
plot(pca$x[,1:2])

# add color
plot(pca$x[,1:2], col=my_cols, pch=16)
```

ggplot PCA
```{r}
#Make a new df with PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```


```{r}
p <- ggplot(my_data) +
      aes(x=PC1, y=PC2,
          size=winpercent/100,
          text=rownames(my_data),
          label=rownames(my_data)) +
  geom_point(col=my_cols)

p
```

Using ggrepel
```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

PCA by loading
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
PC1 resolves the variance between fruity candy (positive direction) and chocolate/bars (negative direction). Looking at the PCA plot we can see PC1 separates the fruity candy (pink) from the chocolate bars (dark brown) well. Pluribus also contributes to PC1 in the positive direction and is associated with the fruity candy type as there are many candies in a package as opposed to chocolate bars where this is only 1 candy inside.